---
title: "Is an automatic or manual transmission better for MPG?"
author: "Lukas Nick"
date: "August 24 2014"
output: pdf_document
---

# Executive summary

# Question

1. Is an automatic transmission better for MPG?
2. Quantify the MPG difference between automatic and manual transmissions.
3. Does the answers to 1) and 2) depend on other variables?
  * number of cylinders
  * weight
  * displacementnumber of forward gears
  * ...?
  
  
  
# Data

I try to answer these questions based on the dataset mtcars, which has 32 cars and the following variables:

```{r echo=FALSE}
setwd("~/Documents/ml/datascience/courses/07_RegressionModels/project/")
library(datasets)
data(mtcars)

# Create factors for the categorial variables
cars <- mtcars
cars$am <- factor(mtcars$am, levels=c(0,1), labels=c("automatic", "manual"))
cars$cyl <- factor(mtcars$cyl)
cars$vs <- factor(mtcars$vs, levels=c(0,1), labels=c("s", "v"))
cars$gear <- factor(mtcars$gear)
cars$carb <- factor(mtcars$carb)

length(rownames(cars))
names(cars)
```


## Results

### Exploratory results

Let's check how the variables correlate with mpg:
```{r echo=FALSE}
cors <- cor(mtcars$mpg,mtcars[,-1])
cor.tests <- sapply(mtcars[, -1], cor.test, mtcars$mpg)
alphas <- sapply(0:9, function(i){ cor.tests[i*9+3]} )

results <- rbind(cors, alphas)
rownames(results) <- c("corr", "p-value")
results
```

Since we want to know the relationship between transmission and mpg, let's check the corrations between the variables and transmission:
```{r}
other.vars <- c(2:8, 10, 11)
cors <- cor(mtcars$am,mtcars[,other.vars])
cor.tests <- sapply(mtcars[,other.vars], cor.test, mtcars$am)
alphas <- sapply(0:9, function(i){ cor.tests[i*9+3]} )

results <- rbind(cors, alphas)
rownames(results) <- c("corr", "p-value")
results
```

Since - in theory - the number of cylinder might have an influence on the relationship between transmission and mpg, plot mpg for 4, 6, and 8 cylinders separately and color the data points by transmission type:
```{r}
par(mfrow=c(1,2))
plot(cars$cyl, cars$mpg, ylab="mpg", xlab="cylinders")
points(jitter(as.numeric(cars$cyl), factor=0.5), cars$mpg, col=sapply(cars$am, switch, 'red', 'green'), pch=16)
legend(x="topleft", legend=c('manual', 'automatic'), pch=16, col=c('red', 'green'), cex=0.8, inset=0.03)

plot(cars$gear, cars$mpg, ylab="mpg", xlab="gears")
points(jitter(as.numeric(cars$gear), factor=0.5), cars$mpg, col=sapply(cars$am, switch, 'red', 'green'), pch=16)
legend(x="topleft", legend=c('manual', 'automatic'), pch=16, col=c('red', 'green'), cex=0.8, inset=0.03)

```


In order to find variables that we should adjust for, find out which variables `am` depends on.   

The categorial variables:
```{r }
sapply( c("cyl", "vs", "gear", "carb"), function(cat){ summary( table(cars[,cat], cars$am) )} )
```

So the variables `cyl` and `gear` differ, depending on `am`.

Influence of the numerical variables:
```{r}
rbind( sapply( c("disp", "hp", "drat", "wt", "qsec"), function(var){ t.test( cars[,var] ~ cars$am)[3] } ) )
```

So the variables`disp`, `drat`, and `wt` differ depending on `am`.

Let's look at some plots to get a feeling of the data distributions:

Categorical variables $cyl$ and $gear$:
```{r}
layout(matrix(c(1,2),1,2))
plot(cars$cyl, cars$mpg, xlab="cylinders", ylab="mpg")
points(jitter(as.numeric(cars$cyl), factor=0.5), cars$mpg, col=ifelse(cars$am=='manual', 'red', 'black'))
legend(x="topright", legend=c('manual', 'automatic'), pch=16, col=c('red', 'black'), cex=0.8, inset=0.03)

plot(cars$gear, cars$mpg, xlab="gears", ylab="mpg")
points(jitter(as.numeric(cars$gear), factor=0.5), cars$mpg, col=ifelse(cars$am=='manual', 'red', 'black'), pch=16)
legend(x="topright", legend=c('manual', 'automatic'), pch=16, col=c('red', 'black'), cex=0.8, inset=0.03)

```

Numerical variables $disp$, $drat$, and $wt$:
```{r echo=FALSE}
library(ggplot2)

# Multiple plot function
# FROM: http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  require(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
p1 <- qplot(disp, mpg, data=cars, geom=c("point", "smooth"), 
      method="lm", formula=y~x, color=am, 
      main="Regression of mpg on disp", xlab="disp", ylab="mpg") + theme_bw() + theme(legend.position = c(0.8, 0.8))

p2 <- qplot(drat, mpg, data=cars, geom=c("point", "smooth"), 
      method="lm", formula=y~x, color=am, 
      main="Regression of mpg on drat", xlab="drat", ylab="mpg") + theme_bw() + theme(legend.position = c(0.2, 0.8))

p3 <- qplot(wt, mpg, data=cars, geom=c("point", "smooth"), 
      method="lm", formula=y~x, color=am, 
      main="Regression of mpg on weight", xlab="wt", ylab="mpg") + theme_bw() + theme(legend.position = c(0.8, 0.8))

multiplot(p1, p2, p3, cols=2)
```

# Regression results

## Simple linear regression

Above boxplot (transmission vs mpg) suggested that manual transmission has higher mpg values than automatic transmission. 
A simple linear regression of `mpg` on `am` confirms this:
```{r}
fit <- lm( mpg ~ am, data=cars)
summary(fit)[4]
```

The mean of mpg for the manual transmission cars is 7.2 miles higher than for the automatical transmission cars, a significant difference (p=0.0003).

### Confidence intervals

With 95% confidence we can estimate that difference between automatic and manual transmission cars is between 3.65 and 10.85 miles per gallon.
```{r echo=FALSE}
sumCoef <- summary(fit)$coefficients
est.automatic <- sumCoef[1,1] + c(-1, 1) * qt(.975, df = fit$df) * sumCoef[1, 2]
est.diff      <- sumCoef[2,1] + c(-1, 1) * qt(.975, df = fit$df) * sumCoef[2, 2]
coef.tab <- rbind(est.automatic, est.diff)
colnames(coef.tab) <- c("lower bound", "upper bound")
rownames(coef.tab) <- c("mean automatic", "manual - automatic")
coef.tab
```

### Diagnostics for the simple regression.

The residuals of the simple regression seem to be randomly and normally distributed:
```{r echo=FALSE}
par(mfrow=c(1,2))
plot(fit.simple, which=1)
plot(fit.simple, which=2)
```

No data point has a special influence:
```{r}
 range(hatvalues(fit.simple))
[1] 0.05263158 0.07692308
```

## Multiple regression

Car weight is related to both transmission and mpg. Looking at the three plots relating `disp`, `drat`, and `wt` to `mpg`, the difference between automatic and manual transmission seemed to be the greatest in the case of weight. So let's first adjust for weight:
```{r}
fit.wt <- lm(mpg ~ am + wt, data=cars)
summary(fit.wt)
```
This model explains much more variance than the model that includes transmission alone (75% as opposed to 36%).  
Adjusting for weight, we see that difference in transmission disappears.

Let's include the interaction term:
```{r}
fit.am.wt <- lm(mpg ~ am*wt, data=cars)
summary(fit.am.wt)
```

Now we see that there is in fact a difference between automatic and manual transmission: the mpg of both manual and automatic transmission cars goes down the heavier the car, but the mpg go down much steeper for manual than for automatic transmission cars. Still, the mean mpg of manual transmission cars is significantly higher than for automatic transmission cars.
The model with the interaction explains 83% of variance (as opposed to 75% for the model without the interaction term).

Based on the relationships established in the exploratory section, `cyl` have an influence on both `mpg` and transmission. Diagram XX shows a similar relationship between transmission, cylinders, and mpg as between transmission, weight, and mpg.

Including cylinders to the interaction (`mpg ~ am*wt + cyl`) makes a significant, but only small difference: it adds only 5% explained variance.
This small additional value of including cylinders is also confirmed by an anova between the models:
```{r}
fit1 <- lm(mpg ~ am, data=cars)
fit2 <- update(fit1, mpg ~ am + wt, data=cars)
fit3 <- update(fit1, mpg ~ am + wt + am*wt, data=cars)
fit4 <- update(fit1, mpg ~ am + wt + am*wt + cyl, data=cars)
anova(fit1, fit2, fit3, fit4)
```

Cylinder is of course related with weight, so it doesn't explain much more.

Trying out various models with additional variables, the only one that makes a significant difference is `mpg ~ aml + wt + cyl`.
```{r}
summary(fit3)
```


## Is an automatic transmission better for MPG?

## How big is the difference in MPG between automatic and manual transmission?

# Conclusions

02_02, p.20: counts violate assumption of normality, so validity of model is limited. Should take log?


